---
title: "Online Appendix for Heterogeneity in Wildlife Abundance Estimation"
author: "Jeff Laake and Bret Collier"
date: "2023-03-02"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
library(xtable)
library(pander)
```

All of the code used to generate results in the paper are provided here including simulation of homogeneous and heterogeneous data and analysis of these data demonstrating bias due to residual heterogeneity and use of covariates and planted animals to remove heterogeneity.  Additionally the code used for analysis of the wild turkey data and the tables and plot can be recreated with the code contained herein. The Rmarkdown document that created this online appendix is also available online with the turkey.txt file that is needed for the example.

## Homogeneous Simulation Example
The following demonstrates the simulation realization of the homogeneous example and the code to show chi-square contingency test and estimation of the parameters and abundance.
```{r}
#simulation with homogeneous detection probabilities
# set random seed so results are the same as in the paper
set.seed(9039921)
# set parameters
ss=1000
p1=0.4
p2=0.6
# which are seen by first and second observer
p1=as.numeric(runif(ss)<=p1)
p2=as.numeric(runif(ss)<=p2)
# create 2 occasion capture history
all_ch=paste(p1,p2,sep="")
#remove those not seen by either observer
ch_hom=all_ch[all_ch!="00"]

# construct contingency table and show
x=table(p1,p2)
x
# perform chi-square contingency test for independence
chisq.test(x)
# show covariance/correlation/correlation test between detections of observers
cov(p1,p2)
cor(p1,p2)
cor.test(p1,p2)

# compute total number observed and detection probabilities from results
n=(sum(x)-x[1,1])
p_hat=c(x[2,2]/sum(x[,2]),x[2,2]/sum(x[2,]))
p_hat
# compute probability that at least one of the observers detected
p_either=(1-prod(1-p_hat))
p_either
# compute estimate of abundance using n/p_either which is an alternate estimator
# from Lincoln-Petersen and will produce the same value. THis is a more universal estimation
# equation that is common to many forms of abundance estimators including distance sampling
# where the number observed is divided by the sample inclusion probability (overall detection
# probability)
Nhat=n/p_either
Nhat
est_hom=c(n,p_either,Nhat)

# compute probability that observer 2 detects what observer 1 missed
p2_0=x[1,2]/sum(x[1,])
p2_0
# compute probability that observer 2 detects what observer 1 detected (often called a duplicate)
p2_1=x[2,2]/sum(x[2,])
p2_1
# compute overall detection probability for each observer
p1=sum(x[2,])/sum(x)
p1
p2=sum(x[,2])/sum(x)
p2

# construct matrix of capture history probabilities
ctab_hom=matrix(c((1-p1)*(1-p2_0),p1*(1-p2_1),(1-p1)*p2_0,p1*p2_1),nrow=2)
x/sum(x)
ctab_hom

# construct matrix of individual probabilities
x_hom=x
prob_hom=c(p1,p2,p2_0,p2_1)
prob_hom
```

## Heterogeneous Simulation Example
The following demonstrates the simulation realization of the hoterogeneous example and the code to show chi-square contingency test and estimation of the parameters and abundance.
```{r}
#simulation with heterogeneous detection probabilities
# cov is a normal random variable with mean 0 and standard deviation that
# functions as the covariate affecting detection probability
cov=rnorm(ss,0,2)

# set detection probabilities for each of the cov values; cov=0 will give same values
# used in the homogeneous simulation
p1=plogis(log(.4/.6)+cov)
p2=plogis(log(.6/.4)+cov)

# randomly decide which of the ss are detected for each observer
p1=as.numeric(runif(ss)<=p1)
p2=as.numeric(runif(ss)<=p2)
# create 2 occasion capture history
all_ch=paste(p1,p2,sep="")
#remove those not seen by either observer
ch=all_ch[all_ch!="00"]
# store cov values for those detected to be used in model with covariate values
obs_cov=cov[p1>0 | p2>0]

# 50 marked animals that are "planted" in the population for later analysis
# these are the simulated animals that would be known and "tracked" such that we know 
# which are never seen by either observer
marked=sample(1:1000,50)
# add a third dummy occasion which is 1 for the 50 known animals and 0 for all others
p3=rep(0,1000)
p3[marked]=1
# create a three occasion capture history
ch_marked=paste(p1,p2,p3,sep="")
# create a factor variable that has value "marked" for the 50 known and "unmarked" 
# for the observed unknown
marked=factor(p3[ch_marked!="000"],labels=c("unmarked","marked"))
# throw away capture histories that are all "000"; note that marked not seen are "001"
ch_marked=ch_marked[ch_marked!="000"]

# now do the same for 25 marked animals
marked25=sample(1:1000,25)
p3=rep(0,1000)
p3[marked25]=1
ch_marked25=paste(p1,p2,p3,sep="")
marked25=factor(p3[ch_marked25!="000"],labels=c("unmarked","marked"))
ch_marked25=ch_marked25[ch_marked25!="000"]


# construct contingency table and show
x=table(p1,p2)
x
# perform chi-square contingency test for independence
chisq.test(x)
# show covariance/correlation/correlation test between detections of observers
cov(p1,p2)
cor(p1,p2)
cor.test(p1,p2)

# Compute detection probabilities and abundance estimate as done above
# assuming independence with the heterogeneous simulation
n=(sum(x)-x[1,1])
p_hat=c(x[2,2]/sum(x[,2]),x[2,2]/sum(x[2,]))
p_hat
p_either=(1-prod(1-p_hat))
p_either
p_neither=prod(1-p_hat)
p_neither

Nhat=n/p_either
Nhat

est_het=c(n,p_either,Nhat)

# compute probability that observer 2 detects what observer 1 missed
p2_0=x[1,2]/sum(x[1,])
p2_0
# compute probability that observer 2 detects what observer 1 detected 
# (often called a duplicate)
p2_1=x[2,2]/sum(x[2,])
p2_1
# compute overall detection probability for each observer
p1=sum(x[2,])/sum(x)
p1
p2=sum(x[,2])/sum(x)
p2

prob_het=c(p1,p2,p2_0,p2_1)
prob_het

# construct matrix of capture history probabilities under dependence model
ctab_het=matrix(c((1-p1)*(1-p2_0),p1*(1-p2_1),(1-p1)*p2_0,p1*p2_1),nrow=2)
x/sum(x)
ctab_het

x_het=x
```


## Simulation Realization for Homogeneous and Heterogeneous Simulations

The following table shows a single realization of the simulation for both the homogeneity and heterogeneity side-by-side.

```{r,echo=FALSE}
pander(cbind(x_hom,x_het))
```


```{r,echo=FALSE}
prob_table=cbind(prob_hom,prob_het)
rownames(prob_table)=c("$p_1$","$p_2$","$p_{2|0}$","$p_{2|1}$")
colnames(prob_table)=c("Homogeneity","Heterogeneity")
pander(prob_table)
```

## Estimation with Huggins model via MARK/RMark
Here will compute the estimates from above with Huggins model in MARK using the RMark interface for the homogeneous and heterogeneous simulation examples.
```{r}

library(RMark)
# create data frame using the capture history from the homogeneous simulation realization
dhet=data.frame(ch=ch_hom,freq=rep(1,length(ch_hom)))
# process data with Huggins model
dp=process.data(dhet,model="Huggins")
# compute default design data
ddl=make.design.data(dp)
# run model with time variation in detection probability; the share=TRUE means 
# recapture probability (p_2|1) is the same as initial capture probability
mod_hom=mark(dp,ddl=ddl,model.parameters=list(p=list(formula=~time,share=TRUE)))


# create data frame using the capture history from the homogeneous simulation realization
dhet=data.frame(ch=ch,freq=rep(1,length(ch)),cov=obs_cov)
# process data with Huggins model
dp=process.data(dhet,model="Huggins")
# compute default design data
ddl=make.design.data(dp)
# run model with time variation in detection probability; the share=TRUE means 
# recapture probability (p_2|1) is the same as initial capture probability
mod_het=mark(dp,ddl=ddl,model.parameters=list(p=list(formula=~time,share=TRUE)))
# run model with time and covariate variation in detection probability; the share=TRUE 
# means recapture probability p_(2|1) is the same as initial capture probability
mod_cov=mark(dp,ddl=ddl,model.parameters=list(p=list(formula=~time+cov,share=TRUE)))
```

## Estimation with "planted" individuals for heterogeneous simulation data
```{r}

# construct probability table for marked (known) and unmarked(unknown) animals
prob_table=cbind(c("Marked Animals","001","011","101","111"," ",
"Unmarked Animals","010","100","110"),c("","$(1-p_1)(1-p_2)p_3$" ,"$(1-p_1)p_2p_3$",
"$p_1(1-c_2)p_3$", "$p_1c_2p_3$" ,"","","$(1-p_1)p_2(1-p_3)$",
"$p_1(1-c_2)(1-p_3)$","$p_1c_2(1-p_3)$"))

# construct dataframe with capture history(ch) and factor variable marked which 
# is used to create groups in the data for abundance estimation
df=data.frame(ch=ch_marked,marked=marked)
# get the length of the capture history which is 3 in this case. This is used to fix 
# p for last occasion
last_time=nchar(df$ch)[1]
# process data with Huggins model and groups defined by marked factor variable
dp=process.data(df,model="Huggins",groups="marked")
# create default design data
ddl=make.design.data(dp)
# add a field "fix" to the design data for parameter p (capture) and c(recapture) 
# a value of NA means that parameter will be estimated
ddl$p$fix=NA
ddl$c$fix=NA
# for the last occasion (time) fix p and c to be 1 for known animals and 0 for
# unknown animals
ddl$p$fix[ddl$p$time==last_time&ddl$p$marked=="marked"]=1
ddl$p$fix[ddl$p$time==last_time&ddl$p$marked=="unmarked"]=0
ddl$c$fix[ddl$c$time==last_time&ddl$c$marked=="marked"]=1
ddl$c$fix[ddl$c$time==last_time&ddl$c$marked=="unmarked"]=0
# show the design data
ddl
# fit a model with time variation and a time constant difference for 
# recapture probability (+c)
mod=mark(dp,ddl,model.parameters=list(p=list(formula=~time+c,share=TRUE)))
# show the derived abundance estimates for each group. known animals will be the number 
# marked because p for last occasion is 1 (all are "detected")
mod$results$derived

# show the estimates for p1,p2 and recapture probability
p1.est=get.real(mod,"p",se=TRUE)[1,]
p1.est
p2.est=get.real(mod,"p",se=TRUE)[2,]
p2.est
c.est=get.real(mod,"c",se=TRUE)[1,]
c.est

```


Table. Probabilities for each capture history of marked and unmarked animals with a two-sample(double observer) structure and an additional third dummy occasion to incorporate known marked animals. 
```{r,echo=FALSE}
pander(prob_table)
```


```{r}
#Now do the same as above but with a sample of 25 known animals
df=data.frame(ch=ch_marked25,marked=marked25)
last_time=nchar(df$ch)[1]
dp=process.data(df,model="Huggins",groups="marked")
ddl=make.design.data(dp)
ddl$p$fix=NA
ddl$p$fix[ddl$p$time==last_time&ddl$p$marked=="marked"]=1
ddl$p$fix[ddl$p$time==last_time&ddl$p$marked=="unmarked"]=0
ddl$c$fix=NA
ddl$c$fix[ddl$c$time==last_time&ddl$c$marked=="marked"]=1
ddl$c$fix[ddl$c$time==last_time&ddl$c$marked=="unmarked"]=0
mod=mark(dp,ddl,model.parameters=list(p=list(formula=~time+c,share=TRUE)))

p1.est=get.real(mod,"p",se=TRUE)[1,]
p1.est
p2.est=get.real(mod,"p",se=TRUE)[2,]
p2.est
c.est=get.real(mod,"c",se=TRUE)[1,]
c.est

```

## Wild turkey example using "planted" individuals

```{r,turkey}
# read in data from wild turkey example
turkey=import.chdata("turkey.txt")
# turkey$ch is capture history for 3 occasions with each "occasion" 
# being 10 consecutive days
# turkey$ch2 is capture history for 5 occasions with each "occasion" being 
# 6 consecutive days
# turkey$ch3 is capture history for 10 occasions with each "occasion" being 
# 3 consecutive days
turkey$ch2=as.character(turkey$ch2)
turkey$ch3=as.character(turkey$ch3)

# use 3 capture occasions - 10 days each
df=turkey
# define factor variable type (same as marked above) with values known (gps/vhf turkeys)
# and unknown
df$ch[df$type=="unknown"]=paste(df$ch[df$type=="unknown"],"0",sep="")
df$ch[df$type=="known"]=paste(df$ch[df$type=="known"],"1",sep="")
# get the length of the capture history which is 3 in this case. This is used to 
# fix p for last occasion
last_time=nchar(df$ch)[1]
# remove all capture histories with all 0s
df=df[df$ch!=paste(rep("0",last_time),collapse=""),]

# process data with Huggins model and type used to create groups
dp=process.data(df,model="Huggins",groups="type")
# create default design data
ddl=make.design.data(dp)
# fix p an dc for last occasion based on value of type as done with simulation example
ddl$p$fix=NA
ddl$p$fix[ddl$p$time==last_time&ddl$p$type=="known"]=1
ddl$p$fix[ddl$p$time==last_time&ddl$p$type=="unknown"]=0
ddl$c$fix=NA
ddl$c$fix[ddl$c$time==last_time&ddl$c$type=="known"]=1
ddl$c$fix[ddl$c$time==last_time&ddl$c$type=="unknown"]=0

# create a function to fit 6 different possible models
fit_models=function()
{
  # constant detection probability
  p.0=list(formula=~1,share=TRUE)    
  # time variation in detection probability
  p.1=list(formula=~time,share=TRUE)  
  # as p.1 with additive difference for recapture probability
  p.2=list(formula=~time+c,share=TRUE)  
  # time variation in capture and recapture probability (interaction)
  p.3=list(formula=~time+c:time,share=TRUE)
  # model in Griffin et al with(2013) additive type difference
  p.4=list(formula=~time+type,share=TRUE)  
  # same a p.4 but time variation differing between types
  p.5=list(formula=~time*type,share=TRUE)      
  # create model list
  cml=create.model.list("Huggins")  
  # fit each model and construct a model results table
  results=mark.wrapper(cml,data=dp,ddl=ddl)    
  return(results)
}
# fit models and store in results3
results3=fit_models()
results3


# same as above with 5 capture occasions of 6 days each
df=turkey
df$ch=df$ch2
df$ch[df$type=="unknown"]=paste(df$ch[df$type=="unknown"],"0",sep="")
df$ch[df$type=="known"]=paste(df$ch[df$type=="known"],"1",sep="")
last_time=nchar(df$ch)[1]
df=df[df$ch!=paste(rep("0",last_time),collapse=""),]

dp=process.data(df,model="Huggins",groups="type")
ddl=make.design.data(dp)
ddl$p$fix=NA
ddl$p$fix[ddl$p$time==last_time&ddl$p$type=="known"]=1
ddl$p$fix[ddl$p$time==last_time&ddl$p$type=="unknown"]=0
ddl$c$fix=NA
ddl$c$fix[ddl$c$time==last_time&ddl$c$type=="known"]=1
ddl$c$fix[ddl$c$time==last_time&ddl$c$type=="unknown"]=0

results5=fit_models()
results5

# same as above with 10 capture occasions - 3 days each
df=turkey
df$ch=df$ch3
df$ch[df$type=="unknown"]=paste(df$ch[df$type=="unknown"],"0",sep="")
df$ch[df$type=="known"]=paste(df$ch[df$type=="known"],"1",sep="")
last_time=nchar(df$ch)[1]
df=df[df$ch!=paste(rep("0",last_time),collapse=""),]

dp=process.data(df,model="Huggins",groups="type")
ddl=make.design.data(dp)
ddl$p$fix=NA
ddl$p$fix[ddl$p$time==last_time&ddl$p$type=="known"]=1
ddl$p$fix[ddl$p$time==last_time&ddl$p$type=="unknown"]=0
ddl$c$fix=NA
ddl$c$fix[ddl$c$time==last_time&ddl$c$type=="known"]=1
ddl$c$fix[ddl$c$time==last_time&ddl$c$type=="unknown"]=0

results10=fit_models()
results10


# Models in p.4 and p.5 that use an type difference as in Griffin et al. (2013)
# They proposed fitting this model whee known animals would have an expected lower 
# detection probability and then using that detection probability estimate for the
# unknown animals.  MARK does not operate that way with groups. It uses the estimated 
# probabilities for each group with the numbers observed for that group to derive the
# group abundance estimates.  Thus to estimate abundance for models based on p.4 and p.5
# we had to create code to compute the abundance estimates. That code is in the
# function compute_Nhat. It computes the estimate, its std error, and 
# 95% log-normal confidence 
# interval as done in MARK. The arguments for the function are a particular model (mod) and
# the number of occasions (nocc) which determines the number of time specific probabilities.
compute_Nhat=function(mod,nocc)
{
  # get real estimates of p_time 
  real_list=get.real(mod,"p",vcv=TRUE)
  p_est=real_list$estimates$estimate[1:nocc]
  vcv=real_list$vcv[1:nocc,1:nocc]
  pdot=1-prod(1-p_est)
  deriv=1-p_est
  vc_pdot=t(deriv)%*%vcv%*%deriv
  Mt1=colSums(mod$data$freq)[2]
  known=colSums(mod$data$freq)[1]
  Nhat=Mt1/pdot
  deriv=matrix(-Mt1/pdot^2,nrow=1)
  se_Nhat=sqrt(Mt1*((1-pdot)/pdot^2)+deriv%*%vc_pdot%*%t(deriv))
  Nhat=Nhat-Mt1
  cv_Nhat=se_Nhat/Nhat
  C=exp(1.96*sqrt(log(1+cv_Nhat^2)))
  Nhat_lcl=Nhat/C+Mt1+known
  Nhat_ucl=Nhat*C+Mt1+known
  Nhat=Nhat+Mt1+known
  return(list(Nhat=Nhat,se=se_Nhat,Nhat_lcl=Nhat_lcl,Nhat_ucl=Nhat_ucl))
}

# This function uses compute_Nhat above for models p.4 and p.5(5th and 6th models)
# and the derived abundance estimates from MARK output for first 4 models and then 
# computes a model average estimate across the 6 models
do_ma=function(results,nocc)
{
  estimate=matrix(0,nrow=nrow(results$model.table),ncol=2,byrow=TRUE)
  vcv=list()
  for(i in 1:nrow(results$model.table))
  {
     model_num=as.numeric(row.names(results$model.table)[i])
     if(model_num<=4)
     {
      estimate[i,]=results[[model_num]]$results$derived[[1]]$estimate
      vcv[[i]]=
        results[[as.numeric(row.names(results$model.table)[i])]]$results$derived.vcv[[1]]
     }else
     {
       Nlist=compute_Nhat(results[[model_num]],nocc)
       vcvn=matrix(0,nrow=2,ncol=2)
       vcvn[2,2]=Nlist$se^2
       known=colSums(results[[model_num]]$data$freq)[1]
       estimate[i,]=c(known,Nlist$Nhat-known)
       vcv[[i]]=vcvn
      }
  }
  weight=results$model.table$weight
  
  return(model.average(x=list(estimate=estimate,weight=weight,vcv=vcv)))
}

# create a matrix of estimates, std errors and confidence intervals of model
# averaged estimates for 3,5 and 10 occasions
Nmat=matrix(0,nrow=3,ncol=5)
observed=nrow(df)
Nhat3=do_ma(results3,3)
est=sum(Nhat3$estimate)-observed
se=sum(Nhat3$se)
cv=se/est
C=exp(1.96*sqrt(log(1+cv^2)))
Nmat[1,]=c(3,observed+est,se,observed+est/C,observed+est*C)
Nhat5=do_ma(results5,5)
est=sum(Nhat5$estimate)-observed
se=sum(Nhat5$se)
cv=se/est
C=exp(1.96*sqrt(log(1+cv^2)))
Nmat[2,]=c(5,observed+est,se,observed+est/C,observed+est*C)
Nhat10=do_ma(results10,10)
est=sum(Nhat10$estimate)-observed
se=sum(Nhat10$se)
cv=se/est
C=exp(1.96*sqrt(log(1+cv^2)))
Nmat[3,]=c(10,observed+est,se,observed+est/C,observed+est*C)
colnames(Nmat)=c("# of occasions", "$\\hat{N}$","Std. Error", "95% LCL", "95% UCL")
```


Table. Capture histories of known and unknown turkeys for three 10 day occasions from Buckley et al. (in press). Only known turkeys can have a "000" capture history.
```{r,echo=FALSE}
pander(with(turkey[!(turkey$ch=="000" &turkey$type=="unknown"),],table(ch,type)))
```


Table. Model selection table from 3 occasions of 10 days each.
```{r,echo=FALSE}
pander(results3$model.table[,3:8])
```

Table. Model selection table from 5 occasions of 6 days each.
```{r,echo=FALSE}
pander(results5$model.table[,3:8])
```


Table. Model selection table from 10 occasions of 3 days each.
```{r,echo=FALSE}
pander(results10$model.table[,3:8])
```

Table. Model averaged abundance estimates for 3,5,10 occasions.
```{r,echo=FALSE}
pander(Nmat)
```


```{r}
# compute abundance and plot estimates and 95% confidence intervals for each of the 6 models
# for 3 occasion turkey data example
png("Figure1.png")
Nhat=NULL
LCL=NULL
UCL=NULL
for(i in 1:6)
{
  if(i %in%1:4)
  {
      Nhat=c(Nhat,sum(results3[[i]]$results$derived[[1]]$estimate))
      LCL=c(LCL,sum(results3[[i]]$results$derived[[1]]$lcl))  
      UCL=c(UCL,sum(results3[[i]]$results$derived[[1]]$ucl))  
  } else
  {
    Nlist=compute_Nhat(results3[[i]],3)
    Nhat=c(Nhat,Nlist$Nhat)
    LCL=c(LCL,Nlist$Nhat_lcl) 
    UCL=c(UCL,Nlist$Nhat_ucl)
  }
}
plot(1:6,Nhat,ylim=c(40,130),xlab="Model No.",ylab="Abundance estimate",las=1)
for(i in 1:6)
  lines(x=c(i,i),y=c(LCL[i],UCL[i]))
lines(c(1,6),c(80,80))
dev.off()
```

